Variation generation and mean fitness increase: \\ Necessity is the mother of genetic invention

### 2017-11-07 ms.tex
reorganization and other comments by MWF

% Abstract
\begin{abstract}
Generation of variation may be detrimental in well-adapted populations evolving under constant selection.
In a constant environment, genetic modifiers that reduce the rate at which variation is generated by processes such as mutation and migration, succeed.
However, departures from this \emph{reduction principle} have been demonstrated.
Here we analyze a general model of evolution under constant selection where the rate at which variation is generated can depend on the fitness of the individual.
We find a general and clear result:
if a modifier allele increases the rate at which below-average individuals generate variation, then it will increase in frequency and increase the  population mean fitness.
This principle applies to phenomena such as stress-induced mutagenesis, condition-dependent dispersal and outcrossing,
and even to the proverb: \emph{"Necessity is the mother of genetic invention"}.
\end{abstract}

@@

The assumptions and conditions for the validity of the \emph{unified reduction principle} by neutral genetic modifiers are~\citep{Altenberg2017}:
(i) effectively infinite population size, (ii) constant-viability selection, (iii) a population at an equilibrium, and (iv) \emph{linear variation} -- the equal scaling of transition probabilities by the modifier.
A departure from the latter assumption occurs if two variation producing processes interact~\citep{Feldman1980,Altenberg2012}.
Departures from the \emph{reduction principle} have also been demonstrated when conditions (i)-(iii) are not met, see for example~\citet{Holsinger1986} and references therein.

@@

\citet{Ram2012} stated that their result represents a departure from the \emph{reduction principle}, but did not explain this departure.
Their analysis was specific to a model that classified individuals by the number of mutant genes in their genotype, similar to models studied by~\citet{Kimura1966a} and~\citet{Haigh1978}.
Moreover, their argument was based on expected increase of the population mean fitness by alleles that modify the mutation rate, rather than on the invasion success of such modifiers~\citep[i.e., analysis of \emph{evolutionary genetic stability}, see][]{Lessard1990}.

@@

% Model
\section*{The Models}

\paragraph*{General model.}\label{sec:general_model}
Consider a large haploid population with an arbitrary set of types $A_1, A_2, \ldots, A_n$.
The change in the frequencies of these types is described by the transformation $f \to f'$: 
\begin{equation}\label{eq:model}
\bar{w} f' = (I-C+MC)D f,
\end{equation}
where $f_k$ is the frequency of type $A_k$ ($f=(f_1, f_2, \ldots, f_n)$ is a stochastic vector: $\sum_{k=1}^n{f_k} = 1$);
$C_k$ is the probability that an individual of type $A_k$ will transition to another type ($C$ is a diagonal non-negative matrix with diagonal elements $C_k$);
$M_{i,j}$ is the transition probability from type $A_j$ to type $A_i$ given that an individual of type $A_j$ is transitioning ($M$ is a column-stochastic matrix: $\sum_{j=1}^n {M_{i,j}} = 1$ for all \emph{j}); 
$D_k$ is the relative fitness of type $A_k$ ($D$ is a diagonal matrix with diagonal elements $D_k$, where we can assume $0 < D_k \le 1$, i.e., fitness values are relative to the fitness of the fittest type); 
$I$ is the identity matrix;
and $\bar{w}$ is the normalizing factor such that $\sum_{k=1}^n{f'_k}=1$ and is equal to the population mean fitness (see~\nameref{sec:AppA}).

The types $A_k$ can represent a single or multiple genetic loci or non-genetic traits.
Importantly, type transmission is vertical and uni-parental (the type is transmitted from a single parent to the offspring) and is independent of the frequencies of the other types.
Specifically, this model precludes processes such as recombination, social learning, sexual outcrossing, and any horizontal or oblique transmission, as these processes are frequency-dependent~\cite[pg.~54]{Cavalli-Sforza1981}.

Transition between types is determined by a combination of two effects:
(i) the probability to transition \emph{out} of type $A_k$ is determined by $C_k$;
(ii) given a transition out of type $A_k$, the distribution of the destination types $A_j$ is given by $M_{j,k}$ (note the index order).

We focus on a neutral modifier locus in complete linkage to the types $A_k$, with no direct effect on fitness, and whose sole function is to determine $C_k$ the rates of transition out of the different types.
Importantly, this modifier can determine different rates for different types; that is, $C_j \ne C_i$ for some $i,j$. The case $C_j = C_j$ for all $i,j$ is covered by the \emph{reduction principle}~\citep[see][]{Altenberg2017}.

% Modifier model
\paragraph*{Modifier model.}
Consider the case of two modifier alleles, $m$ and $M$, inducing different transition probabilities $C=\diag{C_1, \ldots, C_n}$ and $\tilde{C}=\diag{\tilde{C}_1, \ldots, \tilde{C}_n}$, respectively.
The frequencies of type $A_k$ linked to modifier $m$ or $M$ are $f_k$ and $g_k$, respectively, such that $\sum_{k=1}^{n}{(f_k + g_k)}=1$.
$\bar{w}$ now ensures that $\sum_{k=1}^{n}{(f_k' + g_k')}=1$, 
and the rest of the model parameters are the same as in Eq.~\ref{eq:model}.

The frequencies in the next generation, $f'$ for allele $m$ and $g'$ for allele $M$, are given by
\begin{equation}
\begin{cases}
\bar{w} f' &= (I-C+MC)D f \\
\bar{w} g' &= (I-\tilde{C}+M\tilde{C})D g
\end{cases}.
\label{eq:modifier_model}
\end{equation}
Note that Eq.~\ref{eq:model} is a special case for the absence of allele $M$, with $g_k=0$ for all $k$.
Here, $\bar{w}=\sum_{k=1}^{n}{(f_k+g_k)D_k}$ is the mean fitness of the entire population (see~\nameref{sec:AppA}).

In the following section we present three particular examples of the model presented in Eq.~\ref{eq:model} for the modification of mutation and migration.
For simplicity of the presentation we focus on one modifier allele in these examples.

% Mutation model 1
\paragraph*{Mutation modification model 1.}\label{sec:mut_model_1}

Here we consider a large population of haploids and a trait determined by a major locus with $n$ possible alleles $A_1, A_2, ..., A_n$.
Linked to to this locus is a modifier locus that determines the mutation rates $\mu_k$ of individuals with allele $A_k$.
Specifically, with probability $1-\mu_k$, the allele $A_k$ does not mutate, and with probability $\frac{\mu_k}{n-1}$, the allele $A_k$ mutates to $A_j$ for any $j \ne k$.
This is an extension of a model studied by~\citet{Altenberg2017}, that allows for the mutation rate of an individual, $\mu_k$, to depend on the identity of the allele $A_k$.

@@

This model can be rephrased in terms of the general model: 
Eq.~\ref{eq:mutation_model_1} is equivalent to Eq.~\ref{eq:model} if we define
\begin{equation} 
C_k = \mu_k, \quad
D_k = w_k, \quad
M_{i,j} = \begin{cases}
\frac{1}{n-1}, & i \ne j \\
0, & i=j
\end{cases}.
\end{equation}

% Mutation model 2
\paragraph*{Mutation modification model 2.}\label{sec:mut_model_2}

Here, we consider a large population of haploids in which individuals with genotype $A_k$ are characterized by the number $k$ of deleterious or mutant alleles in their genotype, where $0 \le k \le n$.
A modifier locus then determines the probability $m_k$ that a mutation occurs in individuals with $k$ mutant alleles.
In this case, we assume a low mutation rate $m_k$ so that two mutations are unlikely to occur in the same individual in one generation.
When a mutation occurs it is \emph{deleterious} with probability $\delta$, generating a mutant allele and converting the individual from $A_k$ to $A_{k+1}$, or it is beneficial with probability $\beta$, converting the individual from $A_k$ to $A_{k-1}$.
Note that such beneficial mutations can be either compensatory or back-mutations, and that mutations are neutral with probability $1-\delta-\beta$.
This model has been analyzed by~\citet{Ram2012}.

@@

\paragraph*{Migration modification model.}\label{sec:mig_model_1}

Here we assume a large population of haploids that occupies two demes and a modifier locus that determine the migration rates between the demes, $m_k$. 
Let the fitness of genotypes $A_k$ be $w_k$ in deme 1 and $q_k$ in deme 2 for $k=1,2,\ldots,n$.
Let the frequencies of $A_k$ be $x_k$ in deme 1 and $y_k$ in deme 2 for $k=1,2,\ldots,n$ with $x_k \ge 0$, $y_k \ge 0$, $\sum_{k=1}^{n}{x_k} = 1$, and $\sum_{k=1}^{n}{y_k} = 1$.

Then following selection and migration, after one generation the new frequencies $x_k'$ and $y_k'$ are given by
\begin{align}
x_k' = (1-m_k) \frac{w_k}{\bar{w}} x_k + m_k \frac{q_k}{\bar{q}} y_k \\
y_k' = m_k \frac{w_k}{\bar{w}} x_k + (1 - m_k) \frac{q_k}{\bar{q}} y_k
\end{align}
for $k=1,2,\ldots,n$, where $\bar{w} = \sum_{k \ge 0}{w_k x_k}$ and $\bar{q} = \sum_{k \ge 0}{q_k y_k}$ are the population mean fitnesses in deme 1 and 2, respectively.
This is an extension of a model presented by~\citet{Altenberg2017}, allowing for the migration rate of an individual, $m_k$, to depend on the identity of the allele $A_k$.

@@

@@

The above lemma provides a condition for the effect of changing $C_k$, the probability for transition from $A_k$, on the stable population mean fitness.
Specifically, if $A_k$ individuals have below-average fitness, then a modifier that increases $C_k$ will increase the population mean fitness.

@@

Given $0 < D_k \le 1$, from Eq.~\ref{mfp_intermediate} we can get the following corollary.

\begin{corollary}[Reproductive value principle]
In the notation of Lemma~\ref{lemma:mfp},
\begin{equation}
sign \frac{\partial \bar{w}^*}{\partial C_k} = 
sign (u\tr [M]_k - u_k),
\end{equation}
and therefore increased transition out of type $k$ will increase the stable population mean fitness if the fraction of the population descended from type $k$ is expected to increase due to a transition to another type. 
\end{corollary}

@@

We turn our attention to a modifier allele that concurrently controls the transition out of a subset of types, $K$.
For example, \citet[Appendix~B]{Ram2012} considered individuals that are grouped by the number of their accumulated deleterious mutations, $k$, 
and a modifier allele that regulates the mutation rates in individuals with at least $\pi$ deleterious mutations.
Similarly, if the transition rates are controlled by a threshold rule 
\begin{equation}
C_k = \begin{cases}
\tau, & k \in K \\
1, & k \not\in K
\end{cases},
\end{equation}
where $\tau > 0$, then the sensitivity of the stable population mean fitness to changes in the
rate of transition from types $K$ can be written 
\begin{equation}
\begin{aligned}
\frac{\partial \bar{w}^*}{\partial \tau} = 
u\tr \Big( \sum_{k \in K}{\frac{\partial (I - C + MC)D}{\partial \tau}} \Big) v = \\
\frac{1}{\tau} \sum_{k \in K}{u_k v_k (\bar{w}^* - D_k)},
\end{aligned}
\end{equation}
which can be simplified to
\begin{multline}\label{mfp_sign_threshold}
sign \frac{\partial \bar{w}^*}{\partial \tau} = 
sign \sum_{k \in K}{u_k v_k (\bar{w}^* - D_k)} = \\ 
sign \Big(\bar{w}^* - \frac{\sum_{k \in K}{u_k v_k D_k}}{\sum_{k \in K}{u_k v_k}}\Big), 
\end{multline}
a generalization of Eq.~\ref{mfp_sign}.
Therefore, increased transition from types $K$ will increase the stable population mean fitness if the average fitness of individuals descended from types $K$ is below the stable population mean fitness.

If such a modifier has the same effect on all individuals (i.e., $C_k = \alpha$ for all $k$) then Eq.~\ref{eq:model_equilibrium} $\bar{w}^* v = ((1-\alpha)I + \alpha M)Dv$,
and the \emph{unified reduction principle} is in effect~\cite[eqs. 65, 72]{Altenberg2017}.
Therefore, we can expect modifiers that reduce $\alpha$ to be favored by natural selection.
Indeed, setting $C = \alpha I$ in Eq.~\ref{eq:model_equilibrium} and proceeding as in the above proof, we find
$$
\frac{\partial \bar{w}^*}{\partial \alpha} = 
\bar{w}^* - 1 \le 0,
$$
because $\bar{w}^* = \sum_{k=1}^{n}{v_k D_k}$ (see~\nameref{sec:AppA}), and $D_k \le 1$ for all $k$.

@@

We determined a condition for a modifier allele that controls that transition rate $C_k$ from type $A_k$ to increase the stable population mean fitness.
Could such an allele increase in frequency when initially rare in the population?
To answer this question we analyze the stability of resident modifier allele $m$ with transition rates $C_k$ to the invasion of a modifier allele $M$ with rates $\tilde{C}_k$, as described in Eqs.~\ref{eq:modifier_model}.

The equilibrium of Eqs.~\ref{eq:modifier_model} when modifier allele $M$ is absent from the population is $(v, 0)$, where $v$ is given in Eq.~\ref{eq:model_equilibrium} and $g_k=0$ for all $k$. 
The stability of allele $m$ to invasion by allele $M$ is determined by the leading eigenvalue $\lambda$ of the external stability matrix $\cl_{ex}$ of the equilibrium $(v,0)$, which, in turn, is determined by the Jacobian $\cj$ of Eq.~\ref{eq:modifier_model} evaluated at the equilibrium $(v,0)$, where
\begin{equation}
\cj = \begin{pmatrix}\cl_{in} & 0 \\ 0 & \cl_{ex} \end{pmatrix}.
\end{equation}
The zero block matrices are due to complete linkage between the modifier and the types $A_k$.

$\cl_{ex}$ can be written as
\begin{equation}\label{eq:Lex}
\cl_{ex} = (I - \tilde{C} + M \tilde{C}) D /\bar{w}^*,
\end{equation}
where $\bar{w}^*$ is the stable population mean fitness in the absence of modifier allele $M$ (see~\nameref{sec:AppA}).

\begin{theorem}[Evolution of increased genetic variation]
The sign of the derivative of the leading eigenvalue $\lambda$ of the external stability matrix $\cl_{ex}$, when evaluated at $\tilde{C}_k=C_k$, is
\begin{equation}
sign \frac{\partial \lambda}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
sign(\bar{w}^* - D_k).
\end{equation}
In addition, the leading eigenvalue is $\lambda=1$ when transition rates of modifier alleles $m$ and $M$ are equal ($\tilde{C}_k=C_k$ for all $k$).

Therefore, an initially rare modifier allele $M$ with transition rates slightly different from the resident modifier $m$ can successfully invade the population ($\lambda>1$) if $M$ increases the probability of transition from types with below-average fitness, thereby increasing the stable population mean fitness.
\end{theorem}

\begin{proof}
First, consider the case the $\tilde{C_k} = C_k$ for all $k$.
Substituting $\tilde{C}=C$ in Eq.~\ref{eq:Lex} and multiplying both sides by $\bar{w}^*$,
\begin{equation}
\bar{w}^* L_{ex} = (I-C+MC) D,
\end{equation}
and since $\bar{w}^*$ is the leading eigenvalue of the RHS (\nameref{sec:AppA}),
the leading eigenvalue of $\cl_{ex}$ is $\lambda=1$.

Now, applying Lemma~\ref{lemma:mfp} (eq.~\ref{mfp_sign}) to Eq.~\ref{eq:Lex}, the sign of the derivative of $\lambda$ with respect to $\tilde{C_k}$ is
\begin{equation}
sign \frac{\partial \lambda}{\partial \tilde{C}_k} =
sign\Big(\lambda - \frac{D_k}{\bar{w}^*}\Big),
\end{equation}
and evaluating this expression for $\tilde{C}_k = C_k$), i.e. at $\lambda=1$, gives us
\begin{equation}
sign \frac{\partial \lambda}{\partial \tilde{C}_k}\bigg\rvert_{\lambda=1} = 
sign(\bar{w}^* - D_k),
\end{equation}
since $\bar{w}^*>0$.
This completes the proof.
\end{proof}

@@

We have shown that under constant-viability selection and in an effectively infinite haploid population at mutation-selection or migration-selection equilibrium, the stable population mean fitness increases if below-average individuals increase the rate at which variation is generated. Furthermore, modifier alleles that induce increased generation of variation in such individuals are favored by natural evolution.

We have given several examples of variation-generating processes for which this principle applies -- mainly mutation and migration -- but our model may apply to other processes as well.
For example, the \emph{reduction principle} applies to ecological models of dispersal, and~\citet{Gueijman2013} have demonstrated that even in homogeneous environments, fitness-associated dispersal increases the population mean fitness and is favored by selection over uniform dispersal.
Similarly, if the transmission fidelity of culturally-transmitted traits depends on the type of the type or fitness of the transmitting individual, we expect that our results will hold.

>>>>

% Abstract
\begin{abstract}
Generation of variation may be detrimental in well-adapted populations evolving under constant selection.
In a constant environment, genetic modifiers that reduce the rate at which variation is generated by processes such as mutation and migration, succeed.
However, departures from this \emph{reduction principle} have been demonstrated.
Here we analyze a general model of evolution under constant selection where the rate at which variation is generated can depend on the fitness of the individual.
We find that if a modifier allele increases the rate at which individuals of below-average fitness generate variation, then it will increase in frequency and increase the  population mean fitness.
This principle applies to phenomena such as stress-induced mutagenesis, condition-dependent dispersal and outcrossing,
and demonstrates the proverb: \emph{"Necessity is the mother of genetic invention"}.
\end{abstract}

@@

The assumptions and conditions for the validity of the \emph{unified reduction principle} by neutral genetic modifiers are~\citep{Altenberg2017}:
(i) effectively infinite population size, (ii) constant-viability selection, (iii) a population at an equilibrium, and (iv) \emph{linear variation} -- the equal scaling of transition probabilities by the modifier.
A departure from the latter assumption occurs if two variation-producing processes interact~\citep{Feldman1980,Altenberg2012}.
Departures from the \emph{reduction principle} have also been demonstrated when conditions (i)-(iii) are not met, see for example~\citet{Holsinger1986} and references therein.

@@

\citet{Ram2012} stated that their result represents a departure from the \emph{reduction principle}, but did not explain this departure.
Their analysis was specific to a model that classified individuals by the number of mutant genes in their genotype, similar to models studied by~\citet{Kimura1966a} and~\citet{Haigh1978}.
Moreover, their argument was based on expected increase of the population mean fitness by alleles that modify the mutation rate, rather than on the invasion success of such modifiers~\citep[i.e., analysis of \emph{evolutionary genetic stability}, see][]{Eshel1982, Lessard1990}.

@@

% Model
\section*{Models}

\paragraph*{General model.}\label{sec:general_model}
Consider a large haploid population with an arbitrary set of types $A_1, A_2, \ldots, A_n$.
The change in the frequencies of these types is described by the transformation $f \to f'$: 
\begin{equation}\label{eq:model}
\bar{w} f' = (I-C+MC)D f,
\end{equation}
where $f_k$ is the frequency of type $A_k$: $f=(f_1, f_2, \ldots, f_n)$ is a stochastic vector such that $\sum_{k=1}^n{f_k} = 1$;
$C_k$ is the probability that an individual of type $A_k$ will transition to another type ($C$ is a diagonal non-negative matrix with diagonal elements $C_k$);
$M_{i,j}$ is the transition probability from type $A_j$ to type $A_i$ given that an individual of type $A_j$ is transitioning ($M$ is a column-stochastic matrix: $\sum_{j=1}^n {M_{i,j}} = 1$ for all \emph{j}); 
$D_k$ is the relative fitness of type $A_k$ ($D$ is a diagonal matrix with diagonal elements $D_k$, where we can assume $0 < D_k \le 1$, i.e., fitness values are relative to the fitness of the fittest type); 
$I$ is the identity matrix;
and $\bar{w}$ is the normalizing factor such that $\sum_{k=1}^n{f'_k}=1$ and is equal to the population mean fitness (see~\nameref{sec:AppA}).

The types $A_k$ can represent a single or multiple haploid genetic loci or non-genetic traits.
Importantly, type transmission is vertical and uni-parental (the type is transmitted from a single parent to the offspring) and is independent of the frequencies of the other types.
Specifically, this model precludes processes such as recombination, social learning, sexual outcrossing, and any horizontal or oblique transmission, as these processes are frequency-dependent~\cite[pg.~54]{Cavalli-Sforza1981}.

Transition between types is determined by a combination of two effects:
(i) the probability to transition \emph{out} of type $A_k$ is determined by $C_k$;
(ii) given a transition out of type $A_k$, the distribution of the destination types $A_j$ is given by $M_{j,k}$ (note the index order).
Importantly, different types can have different rates; that is, $C_j \ne C_i$ for some $i,j$. The case $C_j = C_j$ for all $i,j$ is covered by the \emph{reduction principle}~\citep[see][]{Altenberg2017}.

In the following section we present three examples of the model presented in Eq.~\ref{eq:model} for mutation and migration.

% Mutation model 1
\paragraph*{Mutation model 1.}\label{sec:mut_model_1}

Here we consider a large population of haploids and a trait determined by a major locus with $n$ possible alleles $A_1, A_2, ..., A_n$.
The mutation rates $\mu_k$ of individuals with allele $A_k$ are potentially different;
specifically, with probability $1-\mu_k$, the allele $A_k$ does not mutate, and with probability $\frac{\mu_k}{n-1}$, the allele $A_k$ mutates to $A_j$ for any $j \ne k$.
This is an extension of a model studied by~\citet{Altenberg2017} that allows for the mutation rate of an individual, $\mu_k$, to depend on the identity of the allele $A_k$.

@@

This model can be rephrased in terms of the general model: 
Eq.~\ref{eq:mutation_model_1} is equivalent to Eq.~\ref{eq:model} if we set
\begin{equation} 
C_k = \mu_k, \quad
D_k = w_k, \quad
M_{i,j} = \begin{cases}
\frac{1}{n-1}, & i \ne j \\
0, & i=j
\end{cases}.
\end{equation}

% Mutation model 2
\paragraph*{Mutation model 2.}\label{sec:mut_model_2}

Here, we consider a large population of haploids in which individuals with genotype $A_k$ are characterized by the number $k$ of deleterious or mutant alleles in their genotype, where $0 \le k \le n$.
Importantly, the probability $m_k$ that a mutation occurs in individuals with $k$ mutant alleles depends on $k$.
We assume a low mutation rate $m_k$ so that two mutations are unlikely to occur in the same individual in one generation.
When a mutation occurs it is \emph{deleterious} with probability $\delta$, generating a mutant allele and converting the individual from $A_k$ to $A_{k+1}$, or it is beneficial with probability $\beta$, converting the individual from $A_k$ to $A_{k-1}$.
Note that such beneficial mutations can be either compensatory or back-mutations, and that mutations are neutral with probability $1-\delta-\beta$.
This model has been analyzed by~\citet{Ram2012}.

@@

% Migration model
\paragraph*{Migration model.}\label{sec:mig_model_1}

Here we assume a large population of haploids that occupies two demes.
Let the fitness of genotypes $A_k$ be $w_k$ in deme 1 and $q_k$ in deme 2 for $k=1,2,\ldots,n$.
The probability that an individual with genotype $A_k$ migrates between the deme is $m_k$; that is, migration rates depend on the genotypes.
Let the frequencies of $A_k$ be $x_k$ in deme 1 and $y_k$ in deme 2 for $k=1,2,\ldots,n$ with $x_k \ge 0$, $y_k \ge 0$, $\sum_{k=1}^{n}{x_k} = 1$, and $\sum_{k=1}^{n}{y_k} = 1$.

Following selection and migration the new frequencies $x_k'$ and $y_k'$ are given by
\begin{align}
x_k' = (1-m_k) \frac{w_k}{\bar{w}} x_k + m_k \frac{q_k}{\bar{q}} y_k \\
y_k' = m_k \frac{w_k}{\bar{w}} x_k + (1 - m_k) \frac{q_k}{\bar{q}} y_k
\end{align}
for $k=1,2,\ldots,n$, where $\bar{w} = \sum_{k \ge 0}{w_k x_k}$ and $\bar{q} = \sum_{k \ge 0}{q_k y_k}$ are the population mean fitnesses in deme 1 and 2, respectively.
This is an extension of a model presented by~\citet{Altenberg2017}.

@@

\subsection*{Mean fitness principle}

We first focus on the stable population mean fitness.
We will show that if the transition rate from types with below-average fitness increases, then the stable population mean fitness increases, too.

@@

The above lemma provides a condition for the effect of changing $C_k$, the probability for transition from $A_k$, on the stable population mean fitness.
Specifically, if $A_k$ individuals have below-average fitness, then increasing $C_k$ will increase the population mean fitness.

@@

Given $0 < D_k \le 1$, from Eq.~\ref{mfp_intermediate} we have the following corollary.

\begin{corollary}[Reproductive value principle]
In the notation of Lemma~\ref{lemma:mfp},
\begin{equation}
sign \frac{\partial \bar{w}^*}{\partial C_k} = 
sign (u\tr [M]_k - u_k);
\end{equation}
therefore, increased transition from type $k$ will increase the stable population mean fitness if the fraction of the population descended from type $k$ is expected to increase due to a transition to another type. 
\end{corollary}

@@

We turn our attention to the case where transition from a subset of types, $K$, is concurrently controlled.
For example, \citet[Appendix~B]{Ram2012} considered individuals that are grouped by the number of their accumulated deleterious mutations, $k$, 
and a modifier allele that regulates the mutation rates in individuals with at least $\pi$ deleterious mutations.
Similarly, if the transition rates are controlled by a threshold rule 
\begin{equation}
C_k = \begin{cases}
\tau, & k \in K \\
1, & k \not\in K
\end{cases},
\end{equation}
where $\tau > 0$, then the sensitivity of the stable population mean fitness to changes in the
rate of transition from types in $K$ can be written 
\begin{equation}
\begin{aligned}
\frac{\partial \bar{w}^*}{\partial \tau} = 
u\tr \Big( \sum_{k \in K}{\frac{\partial (I - C + MC)D}{\partial \tau}} \Big) v = \\
\frac{1}{\tau} \sum_{k \in K}{u_k v_k (\bar{w}^* - D_k)}.
\end{aligned}
\end{equation}
Then
\begin{multline}\label{mfp_sign_threshold}
sign \frac{\partial \bar{w}^*}{\partial \tau} = 
sign \sum_{k \in K}{u_k v_k (\bar{w}^* - D_k)} = \\ 
sign \Big(\bar{w}^* - \frac{\sum_{k \in K}{u_k v_k D_k}}{\sum_{k \in K}{u_k v_k}}\Big), 
\end{multline}
a generalization of Eq.~\ref{mfp_sign}.
Therefore, increased transition from types in $K$ will increase the stable population mean fitness if the average fitness of individuals descended from types in $K$ is below the stable population mean fitness.

\subsection*{Evolutionary genetic stability}

% Modifier model
We now focus on a neutral modifier locus in complete linkage to the types $A_k$, with no direct effect on fitness, and whose sole function is to determine $C_k$, the rates of transition from the different types.
We will show that modifier alleles that increase the stable population mean fitness in accordance with Lemma~\ref{lemma:mfp} are favored by natural selection.

\paragraph*{Modifier model.}
Consider the case of two modifier alleles, $m$ and $M$, inducing different transition probabilities $C=\diag{C_1, \ldots, C_n}$ and $\tilde{C}=\diag{\tilde{C}_1, \ldots, \tilde{C}_n}$, respectively.
The frequencies of type $A_k$ linked to modifier $m$ or $M$ are $f_k$ and $g_k$, respectively, such that $\sum_{k=1}^{n}{(f_k + g_k)}=1$.
$\bar{w}$ now ensures that $\sum_{k=1}^{n}{(f_k' + g_k')}=1$, 
and the rest of the model parameters are the same as in Eq.~\ref{eq:model}.

The frequencies in the next generation, $f'$ for allele $m$ and $g'$ for allele $M$, are given by
\begin{equation}
\begin{cases}
\bar{w} f' &= (I-C+MC)D f \\
\bar{w} g' &= (I-\tilde{C}+M\tilde{C})D g
\end{cases}.
\label{eq:modifier_model}
\end{equation}
Here, $\bar{w}=\sum_{k=1}^{n}{(f_k+g_k)D_k}$ is the mean fitness of the entire population (see~\nameref{sec:AppA}). 
Note that Eq.~\ref{eq:model} is the special case of Eq.~\ref{eq:modifier_model} where allele $M$ is absent; i.e. $g_k=0$ for all $k$.

@@

We havefound a condition for a modifier allele that controls that transition rate $C_k$ from type $A_k$ to increase the stable population mean fitness (Lemma~\ref{lemma:mfp}).
Could such an allele increase in frequency when initially rare in the population?
To answer this we analyze the stability of resident modifier allele $m$ with transition rates $C_k$ to the invasion of a modifier allele $M$ with rates $\tilde{C}_k$~\ref{eq:modifier_model}~\citep[for an introduction to \emph{evolutionary invasion analysis}, see][ch.~12]{Otto2007}

The equilibrium of Eqs.~\ref{eq:modifier_model} when modifier allele $M$ is absent from the population is $(v, 0)$, where $v$ is given in Eq.~\ref{eq:model_equilibrium} and $g_k=0$ for all $k$. 
The stability of allele $m$ to invasion by allele $M$ is determined by the leading eigenvalue $\lambda_1$ of $\cl_{ex}$ the external stability matrix of the equilibrium $(v,0)$, which, in turn, is determined by the Jacobian $\cj$ of Eq.~\ref{eq:modifier_model} evaluated at the equilibrium $(v,0)$, where
\begin{equation}
\cj = \begin{pmatrix}\cl_{in} & 0 \\ 0 & \cl_{ex} \end{pmatrix},
\end{equation}
and $\cl_{in}$ is the local stability matrix of the equilibrium $(v,0)$.
The zero block matrices are due to the complete linkage between the modifier and the types $A_k$ and to the lack of transition (i.e., mutation) between the modifier alleles.

$\cl_{ex}$ can be written as
\begin{equation}\label{eq:Lex}
\cl_{ex} = (I - \tilde{C} + M \tilde{C}) D /\bar{w}^*,
\end{equation}
where $\bar{w}^*$ is the stable population mean fitness in the absence of modifier allele $M$ (see~\nameref{sec:AppA}).
Note that $\bar{w}^*$ is not an eigenvalue of $\cl_{ex}$.

\begin{theorem}[Evolution of increased genetic variation]
Let $\lambda_1$ be the The leading eigenvalue of the external stability matrix $\cl_{ex}$.
If the transition rates induced by the modifier alleles $m$ and $M$ are equal, i.e., $\tilde{C}_k=C_k$ for all $k$, then the following applies
\begin{equation}
\begin{aligned}
\lambda_1=1, \\
sign \frac{\partial \lambda_1}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
sign(\bar{w}^* - D_k).
\end{aligned}
\end{equation}

Therefore, an initially rare modifier allele $M$ with transition rates slightly different from the resident allele $m$ can successfully invade the population ($\lambda_1>1$) if $M$ increases the probability of transition from types with below-average fitness, thereby increasing the stable population mean fitness.
\end{theorem}

\begin{proof}
Substituting $\tilde{C}=C$ in Eq.~\ref{eq:Lex} and multiplying both sides by $\bar{w}^*$,
\begin{equation}
\bar{w}^* L_{ex} = (I-C+MC) D,
\end{equation}
and since $\bar{w}^*$ is the leading eigenvalue of the RHS (see Eq.~\ref{eq:model_equilibrium} and \nameref{sec:AppA}),
the leading eigenvalue of $\cl_{ex}$ is $\lambda_1=1$.

Now, applying Lemma~\ref{lemma:mfp} (eq.~\ref{mfp_sign}) to Eq.~\ref{eq:Lex}, the sign of the derivative of $\lambda_1$ with respect to $\tilde{C_k}$ is
\begin{equation}
sign \frac{\partial \lambda_1}{\partial \tilde{C}_k} =
sign\Big(\lambda_1 - \frac{D_k}{\bar{w}^*}\Big).
\end{equation}
Evaluating this expression for $\tilde{C}_k = C_k$, i.e. at $\lambda_1=1$, gives us
\begin{multline}
sign \frac{\partial \lambda_1}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
sign\Big(1 - \frac{D_k}{\bar{w}^*}\Big) =  
sign(\bar{w}^* - D_k),
\end{multline}
since $\bar{w}^*>0$.
This completes the proof.
\end{proof}

Note that if the modifier has the same effect on all individuals (i.e., $C_k = \alpha$ and $\tilde{C}_k = \alpha^*$ for all $k$) then $\cl_{ex}$ becomes
\begin{equation}
\cl_{ex} = ((1 - \alpha^*)I + \alpha^* M) D /\bar{w}^*,
\end{equation}  
and the \emph{unified reduction principle} is in effect~\cite[eqs. 65, 72]{Altenberg2017}.
Therefore, we can expect that with these modifiers, an allele $M$ that reduces the transition rate, i.e. $\alpha^* < \alpha$, will be favored by natural selection.
Indeed, setting $C = \alpha I$ in Eq.~\ref{eq:model_equilibrium} and proceeding as in the above, we find
$$
\frac{\partial \bar{w}^*}{\partial \alpha} = 
\bar{w}^* - 1 \le 0,
$$
because $\bar{w}^* = \sum_{k=1}^{n}{v_k D_k}$ (see~\nameref{sec:AppA}), and $D_k \le 1$ for all $k$.

@@

We have shown that under constant-viability selection and in an effectively infinite haploid population at mutation-selection or migration-selection equilibrium, the stable population mean fitness increases if individuals with below-average fitness increase the rate at which variation is generated. Furthermore, modifier alleles that induce increased generation of variation in such individuals are favored by natural evolution.

We have given several examples of variation-generating processes for which this principle applies -- namely mutation and migration -- but our model may apply to other processes as well.
For example, the \emph{reduction principle} applies to ecological models of dispersal, and~\citet{Gueijman2013} have demonstrated that even in homogeneous environments, fitness-associated dispersal increases the population mean fitness and is favored by selection over uniform dispersal.
Similarly, if the transmission fidelity of culturally-transmitted traits depends on the type of the type or fitness of the transmitting individual, we expect that our results will hold.
